### 2007-01-22 Documentation/Paper2/paper2.tex
*** empty log message ***

\vspace*{-3.cm}
\begin{flushright}
TUM-HEP-YYY/0X\\
%hep-ph/
\end{flushright}

@@

(Alternative: ``GLoBES 3.0: From the simulation of non-standard physics to user-defined systematics'')

(Alternative: ``GLoBES 3.0: Simulation of non-standard physics and multi-source multi-detector
setups'')

@@

\begin{abstract}
We present Version 3.0 of the GLoBES (``General Long Baseline Experiment Simulator'') software, which is
a simulation tool for short- and long-baseline neutrino oscillation experiments. As new features, \GLOBESN\ allows the simulation of experiments with multiple discrete sources and detectors by the 
concept of user-defined systematics.
In addition, the combination with external information, such as from different experiment classes, is 
simplified by the concept of user-defined priors. As far as the probability calculation is concerned, \GLOBES\ 
now provides an interface for the inclusion of non-standard physics without re-compilation of the software. 
The set of experiment descriptions coming with \GLOBES\ has been updated. For example, built-in fluxes are now
provided for the simulation of beta beams.
\end{abstract}

@@

Neutrino oscillations are now established as the leading flavor
transition mechanism for neutrinos in a long history of many experiments, see
\eg , \Ref~\cite{Barger:2003qi}. Future facilities, using accelerator-based
 neutrino beams or nuclear reactors as neutrino sources, are proposed
for precision measurements of the neutrino oscillation parameters. In the
simulation of these experiments, 
the presence of multiple solutions which are intrinsic to
 neutrino oscillation probabilities~\cite{Fogli:1996pv,Burguet-Castell:2001ez,Minakata:2001qm,Barger:2001yr} 
affect the performance.
Thus, optimization strategies are required which maximally exploit 
complementarity between experiments. The \GLOBES\ software package~\cite{Huber:2004ka}
is a  modern experiment  simulation  and analysis tool for
a highly accurate beam and detector simulation. In addition, it provides powerful 
means to analyze correlations and degeneracies, especially for the combination
of several experiments. Compared to a Monte Carlo simulation, which yields a different
result in each run, it simulates the performance of the {\em average} experiment
(see \Ref~\cite{Schwetz:2006md} for a discussion of the meaning of ``average'').
The advantage of such an average prediction is a tremendous performance gain,
which can be used for systematical parameter space scans. In addition, it 
simplifies the direct comparison of experiments.

@@

With the C-library, one can extract the $\Delta \chi^2$ for all defined 
oscillation channels for an experiment or any combination of experiments.
Of course, also low-level information, such as oscillation
probabilities or event rates, can be obtained. \GLOBES\ includes the
simulation of neutrino oscillations in matter with arbitrary matter 
density profiles. In addition, it allows to simulate the matter density
uncertainty (see, \eg, \Refs~\cite{Huber:2002mx,Ohlsson:2003ip}) and to 
extract the precision on the matter density (see, \eg, \Refs~\cite{Winter:2005we,Gandhi:2006gu}).
 As one of the most
advanced features of \GLOBES , it provides the technology to 
project the $\Delta \chi^2$, which is a function of all oscillation
parameters, onto any subspace of parameters by local minimization. 
This approach allows the inclusion of multi-parameter-correlations,
where external constraints (\eg, on the solar parameters) can be imposed, too.
Applications of the projection mechanism include the projections onto 
the $\stheta$-axis and the $\stheta$-$\deltacp$-plane. In addition, 
all oscillation parameters can be kept free to precisely localize 
degenerate solutions.

@@

The probability calculation in \GLOBES\ is based on the diagonalization of the
Hamiltonian in layers of constant matter density using the standard three flavor
scenario of neutrino oscillations. In the flavor basis, we have
\begin{equation}
\mathcal{H}(n_e) = \frac{1}{2E} \, U \,
\left(
\begin{array}{ccc}
0 & 0 & 0 \\
0  & \sdm & 0 \\
0  & 0 & \ldm \\
\end{array}
\right) \,
U^\dagger + 
\left(
\begin{array}{ccc}
 \pm \sqrt{2} G_F n_e & 0 & 0 \\
0 & 0 & 0 \\
0 & 0 & 0 
\end{array}
\right)
\label{equ:ham}
\end{equation}
with $U$ being the mixing matrix (described by four parameters $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, and $\deltacp$), and $n_e$ the constant electron density
in the respective matter density layer. The electron density is related to the matter density $\rho$ by 
$n_e \simeq 0.5 \, \rho/m_N$ with $m_N$ being the nucleon mass. The sign in the second term depends on whether one
uses neutrinos or antineutrinos. For $k$ matter density layers with densities
$n^k_e$ and thicknesses $x_i$, the oscillation probability then evaluates to
\begin{equation}
P(\nu_\alpha \rightarrow \nu_\beta) = \left| \langle \nu_\beta | \mathcal{E}(n^k_e, x_k) \hdots  \mathcal{E}(n^1_e, x_1) | \nu_\alpha \rangle \right|^2
\label{equ:probs}
\end{equation}
with the evolution operators (CORRECT REFERENCE?)
\begin{equation}
  \mathcal{E}(n^k_e, x_k) = \exp \left( -i \mathcal{H}(n^k_e) x_k \right) = \frac{1}{2E} \, \hat{U}(n^k_e) \, \left( \begin{array}{ccc}
e^{-i \, m_1^k \, x_k} & 0 & 0 \\
0 &  e^{-i \, m_2^k \, x_k} & 0 \\
0 & 0 & e^{-i \, m_3^k \, x_k} 
\end{array}
\right) \, \hat{U}(n^k_e)^\dagger \, .
\end{equation}
Here the Hamiltonian $\mathcal{H}(n^k_e)$ is diagonalized by the unitary mixing matrix in matter $\hat{U}(n^k_e)$
with the eigenvalues $m_1^k$, $m_2^k$, and $m_3^k$. The probabilities in \equ{probs} are used for the
event rate computation, which is described in greater detail in \Ref~\cite{Huber:2004ka}. In the next section,
we will demonstrate how systematics is implemented thereafter.

\begin{figure}[t]
\begin{center}
\includegraphics[width=8cm]{reactorNSI}
\end{center}
\mycaption{\label{fig:reactorNSI} Fit in the $\stheta$-$\sigma_E$-plane for the simulated values
$\stheta=\sigma_E=0$ (figure similar to \Ref~\cite{Blennow:2005yk}). Here $\sigma_E$ is an additional non-standard parameter describing the wave packet width of the neutrino oscillation wave packet. The figure is computed with
{\tt example6.c} coming with the \GLOBESN\ distribution (and higher).}
\end{figure}

\GLOBESN\ allows the modification of this Hamiltonian and the whole probability engine without
re-compilation of the \GLOBES\ software. This is implemented
by using pointers to the functions of the standard probability engine, which can be changed by
registering a different probability engine. For example, a user may just copy the standard probability
engine of \GLOBES , modify it, and register it. An important prerequisite for the ability to handle
more than the standard six oscillation parameters (plus matter density) $\boldsymbol{\lambda}=(\theta_{12},\theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm, \rho)$, \ie,
\begin{equation}
\boldsymbol{\lambda}=(\theta_{12},\theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm, \rho, \eta_1, \hdots, \eta_n)
\end{equation}
with $n$ non-standard parameters $\eta_j$. These parameters can be accessed in the usual way, see \Ref~\cite{Manual}.
For example, non-standard Hamiltonian effects in the $e$-$\tau$-sector may be introduced by modifying \equ{ham} to
\begin{equation}
 \mathcal{\hat{H}} = \mathcal{H}(n_e) + 
\left(
\begin{array}{ccc}
0 & 0 & \epsilon_{e \tau} \\
0 & 0 & 0 \\
\epsilon_{e \tau}^* & 0 & 0 
\end{array}
\right)
\label{equ:nsi}
\end{equation}
In this case, there are two more real parameters, such as the absolute value and phase of $\epsilon_{e \tau}$.
In the literature, the non-standard physics feature in \GLOBES\ as experimental feature has been used for damping effects (such as neutrino decoherence, decay, \etc) in \Ref~\cite{Blennow:2005yk}, Hamiltonian-level effects (such as non-standard matter effects) in \Ref~\cite{Blennow:2005qj}, and mass-varying neutrinos in \Ref~\cite{Schwetz:2005fy} by implementing an environment dependence of neutrino mass.  ANY OTHERS (PH!)?
We show in \figu{reactorNSI} a result from \Ref~\cite{Blennow:2005yk} provided as {\tt example6.c} in the \GLOBES\
distribution. Compared to \equ{nsi}, the underlying physics takes place at the probability level, \ie, 
\equ{probs} (for details, see \Ref~\cite{Blennow:2005yk}). In {\tt example6.c}, an analytical implementation
of this physics is used for simplicity.

@@

\GLOBES\ supports four types of systematical errors by default: Signal and background
normalization errors, as well as signal and background tilt or energy calibration errors.
The tilt (T) is implemented as a linear distortion of the spectrum around the center,
and the energy calibration (C) as a distortion (stretching) of the reconstructed energy scale.
For the implementation of these systematical errors, the  ``pull method'' is used~\cite{Fogli:2002pt}, which introduces  so-called nuisance parameters $\zeta_i$. 
For example, for the signal and background normalization errors, the respective event rates 
$s_i$ and $b_i$ in
each bin $i$ are multiplied with\footnote{For details on the event rate calculation, see \Ref~\cite{Huber:2004ka}.}
\begin{equation}
s_i(\zeta_1) := (1+\zeta_1)\cdot s_i \, , \quad b_i(\zeta_2) : =(1+\zeta_2)\cdot b_i \, ,
\end{equation} 
\ie, the rates are scaled by these parameters. The systematics $\chi^2$ is then minimized over
these parameters:
\begin{equation}
\chi^2_\mathrm{pull}(\boldsymbol{\lambda}):=\min_{\{\zeta_i\} } \,\, \left( \,
\chi^2(\boldsymbol{\lambda},
\zeta_1, \ldots, \zeta_k)+ \sum_{j=1}^{k} \frac{\zeta_j^2}{\sigma_{\zeta_j}^2}
\, \right)\,.
\label{equ:chipull}
\end{equation}
Here $\chi^2(\boldsymbol{\lambda}, \zeta_1, \ldots, \zeta_k)$ is the usual Possonian $\chi^2$
depending on the neutrino oscillation parameters $\boldsymbol{\lambda}$ and the nuisance
parameters $\zeta_i$ (which, for instance, scale the signal or background rates).
In addition, Gaussian penalties $\zeta_j^2/\sigma_{\zeta_j}^2$ are added, where 
$\sigma_{\zeta_j}$ corresponds to the actual systematical error.
The core part of \equ{chipull} is the definition of $\chi^2$. For example, for a background-free measurement and
a signal normalization error only, it reads
\begin{equation}
  \chi^2 (\boldsymbol{\lambda},a) =  \sum_{i=1}^{\textrm{\# of bins}}2
     \left( O_{i} - (1 + a) T_{i} +
       (1 + a) T_{i} \log \frac{(1 + a) T_{i}}{O_{i}}  \right) + \frac{a^2}{\sigma_a^2} 
\label{equ:chistandard}
\end{equation}
(JK/PH: CHECK THIS EQUATION IF THIS IS THE ACTUAL IMPLEMENTATION. HERE THE SYSTEMATICS IS\\
IMPLEMENTED IN THE SIMULATED RATES. IN ADDITION, CHECK IF THEORETICAL AND EXP. RATES
ARE CORRECTLY GIVEN. [WW])

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=9cm]{reactor}
\end{center}
\mycaption{\label{fig:reactor} Luminosity scaling of the $\stheta$ sensitivity for different systematical error assumptions (90\% confidence level, figure similar to \Ref~\cite{Huber:2006vr}). The thick curve corresponds to \equ{chi2reactor} for $\sigma_R=\sigma_{\mathrm{corr}}=2.8\%$, $\sigma_N=\sigma_F=\sigma_{\mathrm{uncorr}}=0.6\%$ and an additional calibration error. The figure is computed with {\tt example5.c} provided with the \GLOBES\ software.
 MISLEADING: WHY IS THE LUMINOSITY
OF THE ``FAR DETECTOR'' ON THE X-AXIS? SHOULD THIS NOT BE THE SAME AS THE ONE FOR THE NEAR DETECTOR? 
AT LEAST IF I INTERPRET THE CODE CORRECTLY [WW]}
\end{figure}

Using \GLOBESN\ and higher, one can modify the standard $\chi^2$ function in
\GLOBES\ which corresponds to \equ{chistandard}. In addition, the systematics
concept of ``error dimensions'' is replaced by a concept directly related to
this $\chi^2$ function. As an example, consider a reactor experiment with near ($N$)
and far ($F$) detectors which are described by a reactor flux uncertainty $\sigma_R$
and two fiducial volume errors of the near $\sigma_N$ and far $\sigma_F$ detectors.
Then \equ{chistandard} will be replaced by
\begin{align}
  \chi^2(\boldsymbol{\lambda},a_N,a_F,a_R) = & \sum_{i=1}^{\textrm{\# of bins}} \sum_{d = N,F} 2
     \left( O_{d,i} - (1 + a_R + a_d) T_{d,i} + \right. \nonumber \\
& \left. + (1 + a_R + a_d) T_{d,i} \log \frac{(1 + a_R + a_d) T_{d,i}}{O_{d,i}}  \right) 
          + \frac{a_R^2}{\sigma_R^2} + \frac{a_N^2}{\sigma_N^2} + \frac{a_F^2}{\sigma_F^2}
\label{equ:chi2reactor}
\end{align}
HOW CAN ONE MAKE THE LEFT BRACKET LARGER? [WW]
This function can be defined with user-defined systematics. Similarly, calibration errors, shape
errors, uncorrelated bin-to-bin errors, \etc , can be introduced. We show in \figu{reactor} the result of
such a calculation, where the thick curve corresponds to \equ{chi2reactor} plus an additional 
energy calibration error.
Furthermore, more complicated setups using multiple discrete sources and detectors can be simulated. For details, we refer to the manual~\cite{Manual}.
In the past, user-defined systematics has (as experimental feature) been used in \Ref~\cite{Huber:2006vr} to
simulate Double Chooz and in \Ref~\cite{Barger:2006kp} to simulate T2KK. ANYTHING ELSE? [WW]

In summary, \GLOBES\ currently supports the following systematics $\chi^2$ functions:
\begin{description}
\item[{\tt chiSpectrumTilt}]
 Spectral information, signal and background normalization errors and tilts 
\item[{\tt chiNoSysSpectrum}]
 No systematical errors, but spectral information
\item[{\tt chiTotalRatesTilt}]
 Total rates only, signal and background normalization errors and tilts 
\item[{\tt chiSpectrumOnly}]
 Spectrum only, no other systematics 
\item[{\tt chiNoSysTotalRates}]
 Total rates only, no systematics
\item[{\tt chiSpectrumCalib}]
 Spectral information, signal and background normalization errors and energy calibration errors 
\item[{\tt chiZero}]
 Passive systematics (returns $\chi^2=0$, only provides the event rates for the access by other
rules or experiments)
\item[{\tt Any other name}]
 User-defined systematics
\end{description}

@@

In order to include correlations and degeneracies, \GLOBES\ provides
(local) marginalization routines to project the fit manifold onto
a subspace. For example, if one wants to compute the $\stheta$ precision,
one scans the $\stheta$ direction and marginalizes for each fixed $\stheta$
over the other oscillation parameters. This marginalization is performed
after the systematics $\chi^2_{\mathrm{pull}}$ in \equ{chipull} (or similar) 
has been determined:
\begin{equation}
 \chi^2_{\mathrm{proj}}(\boldsymbol{\lambda_1}) := \min_{\{ \boldsymbol{\lambda_2} \} } \left( \, \chi^2_{\mathrm{pull}} (\boldsymbol{\lambda_1} , \boldsymbol{\lambda_2} ) \, \right) \, .
\label{equ:margin}
\end{equation}
In this case, the parameter space of oscillation parameters 
$\boldsymbol{\lambda}$ is projected onto the subspace $\boldsymbol{\lambda_1}$.
If one wants to include external input, which only depends on the oscillation
parameters and not on systematics, one can do so at this level. \GLOBES\ provides
the pre-defined possibility to include external Gaussian constraints on the 
oscillation parameters (and the matter density, which is treated as another
oscillation parameter). For example, an external constraint on $\theta_{12}$
is added before the marginalization in \equ{margin} by the replacement
\begin{equation}
 \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda}) \rightarrow \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda})  +
 \frac{(\theta_{12} - \theta_{12}^0)^2}{\sigma_{\theta_{12}}^2}
\label{equ:priors}
\end{equation}
with the central value $\theta_{12}^0$, where the input is added, and
the $1\sigma$ Gaussian error $\sigma_{\theta_{12}}$. This corresponds to an external
measurement of the solar mixing angle at the best-fit value $\theta_{12}^0$
with the error $\sigma_{\theta_{12}}$.

In practice, the external error on $\theta_{12}$ may not be Gaussian or one may
want to add other external information. Therefore,
\GLOBESN\ and higher provides the concept of user-defined priors  $f(\boldsymbol{\lambda})$. 
Instead of \equ{priors}, we then have
\begin{equation}
 \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda}) \rightarrow \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda})  + f(\boldsymbol{\lambda}) \, .
\label{equ:userpriors}
\end{equation}
One can easily imagine that this simple concept allows for a high degree of freedom. Examples where
this feature has been used as an experimental feature are the combination of terrestrial neutrino
data with the information from neutrino telescopes~\cite{Winter:2006ce} and the combination of long-baseline data with atmospheric neutrino data~\cite{Huber:2005ep,Campagne:2006yx} (CORRECT?).
Another very interesting option is the use of penalties in degeneracy localization, as suggested by Thomas
Schwetz. For example, if the minimizer runs into the wrong octant, user-defined priors can be
used to add a penalty and prevent it from doing so.

>>>>

\vspace*{-3.cm}
\begin{flushright}
TUM-HEP-656/07\\
%hep-ph/
\end{flushright}

@@

@@

\begin{abstract}
We present Version 3.0 of the GLoBES (``General Long Baseline Experiment Simulator'') software, which is
a simulation tool for short- and long-baseline neutrino oscillation experiments. As new features, \GLOBESN\ allows for user-defined systematical errors, which can also be used to simulate experiments with
multiple discrete sources and detectors.
In addition, the combination with external information, such as from different experiment classes, is 
simplified. As far as the probability calculation is concerned, \GLOBES\ 
now provides an interface for the inclusion of non-standard physics without re-compilation of the software. 
The set of experiment prototypes coming with \GLOBES\ has been updated. For example, built-in fluxes are now
provided for the simulation of beta beams.
\end{abstract}

@@

Neutrino oscillations are now established as the leading flavor
transition mechanism for neutrinos in a long history of many experiments, see
\eg\ \Ref~\cite{Barger:2003qi} and references therein. Future facilities, using accelerator-based
 neutrino beams or nuclear reactors as neutrino sources, are proposed
for precision measurements of the neutrino oscillation parameters. In the
simulation of these experiments, 
the presence of multiple solutions which are intrinsic to
 neutrino oscillation probabilities~\cite{Fogli:1996pv,Burguet-Castell:2001ez,Minakata:2001qm,Barger:2001yr} 
affect the performance.
Thus, optimization strategies are required which maximally exploit 
complementarity between experiments. The \GLOBES\ software package~\cite{Huber:2004ka}
is a  modern experiment  simulation  and analysis tool for
a highly accurate beam and detector simulation. In addition, it provides powerful 
means to analyze correlations and degeneracies, especially for the combination
of several experiments. Compared to a Monte Carlo simulation, which yields a different
result in each run, it simulates the performance of the {\em average} experiment
(see \Ref~\cite{Schwetz:2006md} for a discussion of the meaning of ``average'').
The advantage of such an average prediction is a tremendous performance gain,
which can be used for systematical parameter space scans. In addition, it 
simplifies the direct comparison of experiments.

@@

With the C-library, one can extract the $\Delta \chi^2$ for all defined 
oscillation channels for an experiment or any combination of experiments.
Of course, also low-level information, such as oscillation
probabilities or event rates, can be obtained. \GLOBES\ includes the
simulation of neutrino oscillations in matter with arbitrary matter 
density profiles. In addition, it allows to simulate the matter density
uncertainty (see, \eg, \Refs~\cite{Huber:2002mx,Ohlsson:2003ip}) and to 
extract the precision on the matter density (see, \eg, \Refs~\cite{Winter:2005we,Gandhi:2006gu}).
 As one of the most
advanced features of \GLOBES , it provides the technology to 
project the $\Delta \chi^2$, which is a function of all oscillation
parameters, onto any subspace of parameters by local minimization. 
This approach allows the inclusion of multi-parameter-correlations,
where external constraints (\eg, on the solar parameters) can be imposed, too.
Applications of the projection mechanism include the projections onto 
the $\stheta$-axis and the $\stheta$-$\deltacp$-plane. In addition, 
all oscillation parameters can be kept free to numerically localize 
degenerate solutions.

@@

The probability calculation in \GLOBES\ is based on the diagonalization of the
Hamiltonian in layers of constant matter density using the standard three flavor
scenario of neutrino oscillations. In the flavor base, we have
\begin{equation}
\mathcal{H}(n_e) = \frac{1}{2E} \, U \,
\left(
\begin{array}{ccc}
0 & 0 & 0 \\
0  & \sdm & 0 \\
0  & 0 & \ldm \\
\end{array}
\right) \,
U^\dagger + 
\left(
\begin{array}{ccc}
 \pm \sqrt{2} G_F n_e & 0 & 0 \\
0 & 0 & 0 \\
0 & 0 & 0 
\end{array}
\right)
\label{equ:ham}
\end{equation}
with $U$ being the mixing matrix (described by four parameters $\theta_{12}$, $\theta_{13}$, $\theta_{23}$, and $\deltacp$), and $n_e$ the constant electron density
in the respective matter density layer. The electron density is related to the matter density $\rho$ by 
$n_e \simeq 0.5 \, \rho/m_N$ with $m_N$ being the nucleon mass. The sign in the second term depends on whether one
uses neutrinos or antineutrinos. For $k$ matter density layers with densities
$n^k_e$ and thicknesses $x_i$, the oscillation probability then evaluates to
\begin{equation}
P(\nu_\alpha \rightarrow \nu_\beta) = \left| \langle \nu_\beta | \mathcal{E}(n^k_e, x_k) \hdots  \mathcal{E}(n^1_e, x_1) | \nu_\alpha \rangle \right|^2
\label{equ:probs}
\end{equation}
with the evolution operators (see, \eg, \Ref~\cite{Ohlsson:1999um})
\begin{equation}
  \mathcal{E}(n^k_e, x_k) = \exp \left( -i \mathcal{H}(n^k_e) x_k \right) = \frac{1}{2E} \, \hat{U}(n^k_e) \, \left( \begin{array}{ccc}
e^{-i \, m_1^k \, x_k} & 0 & 0 \\
0 &  e^{-i \, m_2^k \, x_k} & 0 \\
0 & 0 & e^{-i \, m_3^k \, x_k} 
\end{array}
\right) \, \hat{U}(n^k_e)^\dagger \, .
\end{equation}
Here the Hamiltonian $\mathcal{H}(n^k_e)$ is diagonalized by the unitary mixing matrix in matter $\hat{U}(n^k_e)$
with the eigenvalues $m_1^k$, $m_2^k$, and $m_3^k$. The probabilities in \equ{probs} are used for the
event rate computation, which is described in greater detail in \Ref~\cite{Huber:2004ka}. In the next section,
we will demonstrate how systematics is implemented.

\begin{figure}[t]
\begin{center}
\includegraphics[width=8cm]{reactorNSI}
\end{center}
\mycaption{\label{fig:reactorNSI} Fit in the $\stheta$-$\sigma_E$-plane for the simulated values
$\stheta=\sigma_E=0$ (figure similar to \Ref~\cite{Blennow:2005yk}). Here $\sigma_E$ is an additional non-standard parameter describing the wave packet width of the neutrino oscillation wave packet. If this parameter is
kept free (vertical axis), the $\stheta$ sensitivity (horizontal axis) will be affected. Note that
the standard $\stheta$ sensitivity is given by the curves for $\sigma_E \equiv 0$.
The figure is computed with
{\tt example6.c} coming with the \GLOBESN\ distribution (and higher).}
\end{figure}

\GLOBESN\ allows the modification of this Hamiltonian and the whole probability engine without
re-compilation of the \GLOBES\ software. This is implemented
by using pointers to the functions of the standard probability engine, which can be changed by
registering a different probability engine. For example, a user may just copy the standard probability
engine of \GLOBES , modify it, and register it. An important prerequisite is the ability to handle
more than the standard six oscillation parameters (plus matter density) $\boldsymbol{\lambda}=(\theta_{12},\theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm, \rho)$, \ie,
\begin{equation}
\boldsymbol{\lambda}=(\theta_{12},\theta_{13}, \theta_{23}, \deltacp, \sdm, \ldm, \rho, \eta_1, \hdots, \eta_n)
\end{equation}
with $n$ non-standard parameters $\eta_j$. These parameters can be accessed in the usual way, see \Ref~\cite{Manual}.
For example, non-standard Hamiltonian effects in the $e$-$\tau$-sector may be introduced by modifying \equ{ham} to
\begin{equation}
 \mathcal{\hat{H}} = \mathcal{H}(n_e) + 
\left(
\begin{array}{ccc}
0 & 0 & \epsilon_{e \tau} \\
0 & 0 & 0 \\
\epsilon_{e \tau}^* & 0 & 0 
\end{array}
\right)
\label{equ:nsi}
\end{equation}
In this case, there are two more real parameters, such as the absolute value and phase of $\epsilon_{e \tau}$.
In the literature, the non-standard physics feature in \GLOBES\ as experimental feature has been used for damping effects (such as neutrino decoherence, decay, \etc) in \Ref~\cite{Blennow:2005yk}, Hamiltonian-level effects (such as non-standard matter effects) in \Ref~\cite{Blennow:2005qj}, and mass-varying neutrinos in \Ref~\cite{Schwetz:2005fy} by implementing an environment dependence of neutrino mass. 
We show in \figu{reactorNSI} a result from \Ref~\cite{Blennow:2005yk} provided as {\tt example6.c} in the \GLOBES\
distribution. In this case, the non-standard physics is a loss of coherence in neutrino oscillations
described by an intrinsic width of the mass eigenstate wave packets $\sigma_E$ (in energy space).
Compared to \equ{nsi}, the underlying physics takes place at the probability level, \ie, 
\equ{probs}, and it can be easily implemented analytically (for details, see \Ref~\cite{Blennow:2005yk}). 
In {\tt example6.c}, an analytical implementation
of this physics is used for simplicity.

@@

\GLOBES\ supports four types of systematical errors by default: Signal and background
normalization errors, as well as signal and background tilt or energy calibration errors.
The tilt (T) is implemented as a linear distortion of the spectrum around the center,
and the energy calibration (C) as a distortion (stretching) of the reconstructed energy scale.
For the implementation of these systematical errors, the  ``pull method'' is used~\cite{Fogli:2002pt}, which introduces  so-called nuisance parameters $\zeta_i$. 
For example, for the signal and background normalization errors, the respective event rates 
$s_i$ and $b_i$ in
each bin $i$ are multiplied with\footnote{For details on the event rate calculation, see \Ref~\cite{Huber:2004ka}.}
\begin{equation}
s_i(\zeta_1) := (1+\zeta_1)\cdot s_i \, , \quad b_i(\zeta_2) : =(1+\zeta_2)\cdot b_i \, ,
\end{equation} 
\ie, the rates are scaled by these parameters. The systematics $\chi^2$ is then minimized over
these parameters:
\begin{equation}
\chi^2_\mathrm{pull}(\boldsymbol{\lambda}):=\min_{\{\zeta_i\} } \,\, \left( \,
\chi^2(\boldsymbol{\lambda},
\zeta_1, \ldots, \zeta_k)+ \sum_{j=1}^{k} \frac{\zeta_j^2}{\sigma_{\zeta_j}^2}
\, \right)\,.
\label{equ:chipull}
\end{equation}
Here $\chi^2(\boldsymbol{\lambda}, \zeta_1, \ldots, \zeta_k)$ is the usual Possonian $\chi^2$
depending on the neutrino oscillation parameters $\boldsymbol{\lambda}$ and the nuisance
parameters $\zeta_i$ (which, for instance, scale the signal or background rates).
In addition, Gaussian penalties $\zeta_j^2/\sigma_{\zeta_j}^2$ are added, where 
$\sigma_{\zeta_j}$ corresponds to the actual systematical error.
The core part of \equ{chipull} is the definition of $\chi^2$. For example, for a background-free measurement and
a signal normalization error only, it reads
\begin{equation}
  \chi^2 (\boldsymbol{\lambda},a) =  \sum_{i=1}^{\textrm{\# of bins}}2
     \left( O_{i} - (1 + a) T_{i} +
       (1 + a) T_{i} \log \frac{(1 + a) T_{i}}{O_{i}}  \right) + \frac{a^2}{\sigma_a^2} 
\label{equ:chistandard}
\end{equation}

@@

\begin{figure}[t]
\begin{center}
\includegraphics[width=9cm]{reactor}
\end{center}
\mycaption{\label{fig:reactor} Luminosity scaling of the $\stheta$ sensitivity for different systematical error assumptions (90\% confidence level, figure similar to \Ref~\cite{Huber:2006vr}). The thick curve corresponds to \equ{chi2reactor} for $\sigma_R=\sigma_{\mathrm{corr}}=2.8\%$, $\sigma_N=\sigma_F=\sigma_{\mathrm{uncorr}}=0.6\%$ and an additional calibration error. The figure is computed with {\tt example5.c} provided with the \GLOBES\ software.}
\end{figure}

Using \GLOBESN\ and higher, one can modify the standard $\chi^2$ function in
\GLOBES\ which corresponds to \equ{chistandard}. In addition, the systematics
concept of ``error dimensions'' is replaced by a concept directly related to
this $\chi^2$ function. As an example, consider a reactor experiment with near ($N$)
and far ($F$) detectors which are described by a reactor flux uncertainty $\sigma_R$
and two fiducial volume errors of the near $\sigma_N$ and far $\sigma_F$ detectors.
Then \equ{chistandard} will be replaced by
\begin{align}
  \chi^2(\boldsymbol{\lambda},a_N,a_F,a_R) = & \sum_{i=1}^{\textrm{\# of bins}} \sum_{d = N,F} 2
     \Big( O_{d,i} - (1 + a_R + a_d) T_{d,i} +  \nonumber \\
& \left. + (1 + a_R + a_d) T_{d,i} \log \frac{(1 + a_R + a_d) T_{d,i}}{O_{d,i}}  \right) 
          + \frac{a_R^2}{\sigma_R^2} + \frac{a_N^2}{\sigma_N^2} + \frac{a_F^2}{\sigma_F^2}
\label{equ:chi2reactor}
\end{align}
This function can be defined with user-defined systematics. Similarly, calibration errors, shape
errors, uncorrelated bin-to-bin errors, \etc , can be introduced. We show in \figu{reactor} the result of
such a calculation, where the thick curve corresponds to \equ{chi2reactor} plus an additional 
energy calibration error.\footnote{In fact, \figu{reactor} was computed with Gaussian statistics for
efficiency reasons, whereas \GLOBES\ internally uses Poissonian statistics as a standard; see \Ref~\cite{Manual}.}
Furthermore, more complicated setups using multiple discrete sources and detectors can be simulated. For details, we refer to the manual~\cite{Manual}.
In the past, user-defined systematics has (as experimental feature) been used in \Ref~\cite{Huber:2006vr} to
simulate Double Chooz, in \Ref~\cite{Barger:2006kp} to simulate T2KK, and in
\Ref~\cite{Kopp:2006mw} to simulate various reactor experiment setups and complicated
 backgrounds from geoneutrinos.

In summary, \GLOBES\ currently supports the following systematics $\chi^2$ functions:
\begin{description}
\item[{\tt chiSpectrumTilt}]
 Spectral information, signal and background normalization errors and tilts 
\item[{\tt chiNoSysSpectrum}]
 No systematical errors, but spectral information
\item[{\tt chiTotalRatesTilt}]
 Total rates only, signal and background normalization errors and tilts 
\item[{\tt chiSpectrumOnly}]
 Spectrum only (free signal normalization), no other systematics 
\item[{\tt chiNoSysTotalRates}]
 Total rates only, no systematics
\item[{\tt chiSpectrumCalib}]
 Spectral information, signal and background normalization errors and energy calibration errors 
\item[{\tt chiZero}]
 Passive systematics (returns $\chi^2=0$, only provides the event rates for the access by other
rules or experiments)
\item[{\tt Any other name}]
 User-defined systematics
\end{description}

@@

In order to include correlations and degeneracies, \GLOBES\ provides
(local) marginalization routines to project the fit manifold onto
a subspace. For example, if one wants to compute the $\stheta$ precision,
one scans the $\stheta$ direction and marginalizes for each fixed $\stheta$
over the other oscillation parameters. This marginalization is performed
after the systematics $\chi^2_{\mathrm{pull}}$ in \equ{chipull} (or similar) 
has been determined:\footnote{The marginalization order systematics - correlations
is, strictly speaking, not valid anymore for the new hybrid minimizer provided
with \GLOBESN\ as an experimental feature. However, we keep this description
for pedagogical reasons.}
\begin{equation}
 \chi^2_{\mathrm{proj}}(\boldsymbol{\lambda_1}) := \min_{\{ \boldsymbol{\lambda_2} \} } \left( \, \chi^2_{\mathrm{pull}} (\boldsymbol{\lambda_1} , \boldsymbol{\lambda_2} ) \, \right) \, .
\label{equ:margin}
\end{equation}
In this case, the parameter space of oscillation parameters 
$\boldsymbol{\lambda}$ is projected onto the subspace $\boldsymbol{\lambda_1}$.
If one wants to include external input, which only depends on the oscillation
parameters and not on systematics, one can do so at this level. \GLOBES\ provides
the pre-defined possibility to include external Gaussian constraints on the 
oscillation parameters (and the matter density, which is treated as another
oscillation parameter). For example, an external constraint on $\theta_{12}$
is added before the marginalization in \equ{margin} by the replacement
\begin{equation}
 \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda}) \rightarrow \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda})  +
 \frac{(\theta_{12} - \theta_{12}^0)^2}{\sigma_{\theta_{12}}^2}
\label{equ:priors}
\end{equation}
with the central value $\theta_{12}^0$, where the input is added, and
the $1\sigma$ Gaussian error $\sigma_{\theta_{12}}$. This corresponds to an external
measurement of the solar mixing angle at the best-fit value $\theta_{12}^0$
with the error $\sigma_{\theta_{12}}$.

In practice, the external error on $\theta_{12}$ may not be Gaussian or one may
want to add other external information. Therefore,
\GLOBESN\ and higher provides the concept of user-defined priors  $f(\boldsymbol{\lambda})$. 
Instead of \equ{priors}, we then have
\begin{equation}
 \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda}) \rightarrow \chi^2_{\mathrm{pull}}(\boldsymbol{\lambda})  + f(\boldsymbol{\lambda}) \, .
\label{equ:userpriors}
\end{equation}
One can easily imagine that this simple concept allows for a high degree of freedom. Examples where
this feature has been used as an experimental feature are the combination of terrestrial neutrino
data with the information from neutrino telescopes~\cite{Winter:2006ce} and the combination of long-baseline data with atmospheric neutrino data~\cite{Huber:2005ep,Campagne:2006yx}.
Another very interesting option is the use of penalties in degeneracy localization, as suggested by Thomas
Schwetz. For example, if the minimizer runs into the wrong octant, user-defined priors can be
used to add a penalty and prevent it from doing so.
